Appendix

### 2025-03-13 manuscript_rmarkdown/manuscript_appendix.Rmd
Work on empirical central normality test.

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

The empirical central normality test was derived using data sampled from asymmetric generalised normal distributions, including outliers, to resemble more realistic datasets.
Here we assess the type I error rate of two, less realistic, sets of data:

1. Data sampled from asymmetric generalised normal distributions without outliers.
2. Data sampled from normal distributions without outliers, without any power transformation applied.

Other aspects of the experiment remained the same.
Thus, we first drew $m_d=10000$ random distributions.
For asymmetric generalised normal distributions, each distribution was parametrised with a randomly chosen skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$.
For fully normal distributions, skewness parameter $\alpha = 0.5$ and shape parameter $\beta = 2.0$ were fixed.
Location and scale parameters were set as $\mu = 0$ and $\sigma = 1$, respectively.
$n = \lceil 10^\gamma \rceil$ values were then randomly drawn, with $\gamma \sim U\left(1.47, 3.00\right)$, which led to between $30$ and $1000$ values being drawn to create $\mathbf{X}_i$.
Residuals were then computed after performing robust location- and scale-invariant transformations with the empirical tapered cosine weighting method for the dataset with asymmetric generalised normal distributions,
and without any transformation for the dataset with only normal distributions.

```{r empirical-central-normality-test-appendix, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Type I error rate as function of the test statistic $\\tau_{\\text{ecn}}$ ",
  "for five datasets, with central portion $\\kappa=0.80$. ",
  "The type I error rate is computed from $m_d=10000$ randomly sampled features, ",
  "These are sampled from asymmetric generalized normal distributions, with and without outliers (Box-Cox and Yeo-Johnson), ",
  "or normal distributions with outliers (none). ",
  "The test statistic is computed as the average residual of each feature after (Box-Cox and Yeo-Johnson) robust location- and shift-invariant power transformation, or before (none)."
)

data <- .plot_type_1_error_rate_appendix(plot_theme = plot_theme, manuscript_dir = manuscript_dir)
data$p
```

```{=latex}
\begin{table}
\begin{center}
\caption{Test statistic $\tau_{\text{ecn}}$ for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate for several datasets.}
\label{tab:empirical-central-normality-appendix}
\begin{tabular}{l | c c c c c c c}

\toprule
type I error rate & 0.50 & 0.20 & 0.10 & 0.05 & 0.02 & 0.01 & 0.001 \\

\midrule
Box-Cox                  & 0.047 & 0.073 & 0.090 & 0.106 & 0.126 & 0.140 & 0.188 \\
Box-Cox (no outlier)     & 0.043 & 0.065 & 0.079 & 0.092 & 0.106 & 0.116 & 0.155 \\
Yeo-Johnson              & 0.041 & 0.062 & 0.075 & 0.088 & 0.103 & 0.115 & 0.154 \\
Yeo-Johnson (no outlier) & 0.041 & 0.061 & 0.074 & 0.085 & 0.099 & 0.109 & 0.139 \\
normal distr.            & 0.039 & 0.066 & 0.083 & 0.097 & 0.117 & 0.132 & 0.174 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

The results are shown in Figure \ref{fig:empirical-central-normality-test-appendix} and Table \ref {tab:empirical-central-normality-appendix}.
These indicate that the test behaves similarly for the different datasets.
For low type I error rates, the test statistic proposed in the main manuscript is more conservative than alternatives 
based on residuals after Box-Cox transformations of asymmetric generalised normally distributed features or on residuals from strictly normally distributed features.

>>>>

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

\begin{definition}[empirical central normality test]

Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n\right\} \in \mathbb{R}^n$ be a vector of length $n$,
sorted so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each value in $\mathbf{X}$. For tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group.
Let $z_i = F^{-1}_{\mathcal{N}}(p_i)$ be the corresponding z-score, derived from the quantile function of the normal distribution.

\end{definition}

Critical values for the test statistics of the empirical central normality test 
are found in Table \ref{tab:empirical-central-normality-critical-statistic}.

```{=latex}
\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\text{ecn},\alpha, n}$ for
empirical central normality at $\kappa = 0.70$, as a function of significance 
level $\alpha$ and number of instances $n$. The presented values should be divided by $100$.}
\label{tab:empirical-central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

\toprule
$n$ \textbackslash $\alpha$ & 0.1 & 0.2 & 0.5 & 0.8 & 0.9 & 0.95 & 0.975 & 0.99 & 0.999 \\

\midrule
    5 & 24.02 & 21.26 & 17.37 & 14.02 & 12.04 &  9.57 &  7.64 &  5.54 & 2.35 \\
   10 & 21.58 & 18.53 & 13.82 & 10.23 &  8.75 &  7.65 &  6.79 &  5.87 & 4.23 \\
   20 & 15.12 & 13.17 & 10.03 &  7.74 &  6.80 &  6.13 &  5.58 &  5.04 & 4.07 \\
   50 &  9.99 &  8.70 &  6.69 &  5.26 &  4.67 &  4.26 &  3.92 &  3.59 & 3.02 \\
  100 &  7.13 &  6.21 &  4.83 &  3.79 &  3.38 &  3.08 &  2.87 &  2.63 & 2.24 \\ 
  200 &  5.19 &  4.54 &  3.51 &  2.77 &  2.47 &  2.26 &  2.09 &  1.92 & 1.65 \\ 
  500 &  3.47 &  3.06 &  2.38 &  1.86 &  1.65 &  1.51 &  1.40 &  1.28 & 1.08 \\
 1000 &  2.65 &  2.35 &  1.84 &  1.44 &  1.27 &  1.16 &  1.07 &  0.98 & 0.81 \\
 2000 &  2.13 &  1.91 &  1.53 &  1.20 &  1.06 &  0.96 &  0.88 &  0.79 & 0.65 \\
 5000 &  1.69 &  1.56 &  1.30 &  1.07 &  0.95 &  0.87 &  0.80 &  0.72 & 0.58 \\
10000 &  1.50 &  1.41 &  1.23 &  1.05 &  0.96 &  0.89 &  0.83 &  0.77 & 0.64 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

### 2025-03-13 manuscript_rmarkdown/manuscript_appendix.tex
Work on empirical central normality test.

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-02-18}

@@

The empirical central normality test was derived using data sampled from
asymmetric generalised normal distributions, including outliers, to
resemble more realistic datasets. Here we assess the type I error rate
of two, less realistic, sets of data:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Data sampled from asymmetric generalised normal distributions without
  outliers.
\item
  Data sampled from normal distributions without outliers, without any
  power transformation applied.
\end{enumerate}

Other aspects of the experiment remained the same. Thus, we first drew
\(m_d=10000\) random distributions. For asymmetric generalised normal
distributions, each distribution was parametrised with a randomly chosen
skewness parameter \(\alpha \sim U\left(0.01, 0.99\right)\) and shape
parameter \(\beta \sim U\left(1.00, 5.00 \right)\). For fully normal
distributions, skewness parameter \(\alpha = 0.5\) and shape parameter
\(\beta = 2.0\) were fixed. Location and scale parameters were set as
\(\mu = 0\) and \(\sigma = 1\), respectively.
\(n = \lceil 10^\gamma \rceil\) values were then randomly drawn, with
\(\gamma \sim U\left(1.47, 3.00\right)\), which led to between \(30\)
and \(1000\) values being drawn to create \(\mathbf{X}_i\). Residuals
were then computed after performing robust location- and scale-invariant
transformations with the empirical tapered cosine weighting method for
the dataset with asymmetric generalised normal distributions, and
without any transformation for the dataset with only normal
distributions.

{\centering \includegraphics{manuscript_appendix_files/figure-latex/empirical-central-normality-test-appendix-1}

\caption{Type I error rate as function of the test statistic $\tau_{\text{ecn}}$ for five datasets, with central portion $\kappa=0.80$. The type I error rate is computed from $m_d=10000$ randomly sampled features, These are sampled from asymmetric generalized normal distributions, with and without outliers (Box-Cox and Yeo-Johnson), or normal distributions with outliers (none). The test statistic is computed as the average residual of each feature after (Box-Cox and Yeo-Johnson) robust location- and shift-invariant power transformation, or before (none).}\label{fig:empirical-central-normality-test-appendix}
\end{figure}

\begin{table}
\begin{center}
\caption{Test statistic $\tau_{\text{ecn}}$ for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate for several datasets.}
\label{tab:empirical-central-normality-appendix}
\begin{tabular}{l | c c c c c c c}

\toprule
type I error rate & 0.50 & 0.20 & 0.10 & 0.05 & 0.02 & 0.01 & 0.001 \\

\midrule
Box-Cox                  & 0.047 & 0.073 & 0.090 & 0.106 & 0.126 & 0.140 & 0.188 \\
Box-Cox (no outlier)     & 0.043 & 0.065 & 0.079 & 0.092 & 0.106 & 0.116 & 0.155 \\
Yeo-Johnson              & 0.041 & 0.062 & 0.075 & 0.088 & 0.103 & 0.115 & 0.154 \\
Yeo-Johnson (no outlier) & 0.041 & 0.061 & 0.074 & 0.085 & 0.099 & 0.109 & 0.139 \\
normal distr.            & 0.039 & 0.066 & 0.083 & 0.097 & 0.117 & 0.132 & 0.174 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The results are shown in Figure
\ref{fig:empirical-central-normality-test-appendix} and Table
\ref {tab:empirical-central-normality-appendix}. These indicate that the
test behaves similarly for the different datasets. For low type I error
rates, the test statistic proposed in the main manuscript is more
conservative than alternatives based on residuals after Box-Cox
transformations of asymmetric generalised normally distributed features
or on residuals from strictly normally distributed features.

>>>>

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-03-13}

@@

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\text{ecn},\alpha, n}$ for
empirical central normality at $\kappa = 0.70$, as a function of significance 
level $\alpha$ and number of instances $n$. The presented values should be divided by $100$.}
\label{tab:empirical-central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

\toprule
$n$ \ $\alpha$ & 0.1 & 0.2 & 0.5 & 0.8 & 0.9 & 0.95 & 0.975 & 0.99 & 0.999 \\

\midrule
    5 & 24.02 & 21.26 & 17.37 & 14.02 & 12.04 &  9.57 &  7.64 &  5.54 & 2.35 \\
   10 & 21.58 & 18.53 & 13.82 & 10.23 &  8.75 &  7.65 &  6.79 &  5.87 & 4.23 \\
   20 & 15.12 & 13.17 & 10.03 &  7.74 &  6.80 &  6.13 &  5.58 &  5.04 & 4.07 \\
   50 &  9.99 &  8.70 &  6.69 &  5.26 &  4.67 &  4.26 &  3.92 &  3.59 & 3.02 \\
  100 &  7.13 &  6.21 &  4.83 &  3.79 &  3.38 &  3.08 &  2.87 &  2.63 & 2.24 \\ 
  200 &  5.19 &  4.54 &  3.51 &  2.77 &  2.47 &  2.26 &  2.09 &  1.92 & 1.65 \\ 
  500 &  3.47 &  3.06 &  2.38 &  1.86 &  1.65 &  1.51 &  1.40 &  1.28 & 1.08 \\
 1000 &  2.65 &  2.35 &  1.84 &  1.44 &  1.27 &  1.16 &  1.07 &  0.98 & 0.81 \\
 2000 &  2.13 &  1.91 &  1.53 &  1.20 &  1.06 &  0.96 &  0.88 &  0.79 & 0.65 \\
 5000 &  1.69 &  1.56 &  1.30 &  1.07 &  0.95 &  0.87 &  0.80 &  0.72 & 0.58 \\
10000 &  1.50 &  1.41 &  1.23 &  1.05 &  0.96 &  0.89 &  0.83 &  0.77 & 0.64 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
